<!-- ---------------------------------- -->
# CellTrails Maps
<!-- ---------------------------------- -->
## Graph Layout

*CellTrails* portrays a computed trajectory as collection of trails that can be found in a landscape shaped by individual expression dynamics. To generate such a topographic trail map - the *CellTrails* map - a two-dimensional spatiotemporal ordination of the expression matrix has to be computed. This can be done by any graph layout algorithm using the structural information from the trajectory graph, which is composed of nodes (=samples) and edges (=chronological relation between samples). 
We found that the freely available graph visualization software _yEd_ (http://www.yworks.com/products/yed) has great capabilities to visualize and analyze a trajectory graph. An optimal layout is planar, i.e. it exhibits no crossing edges or overlapping nodes.

Therefore, *CellTrails* enables to export and import the trajectory graph structure using the common _graphml_ file format. This file format can be interpreted by most third-party graph analysis applications, allowing the user to subject the trajectory graph to a wide range of (tree) layout algorithms. In particular, the exported file has additional ygraph attributes best suited to be used with _yEd_, which is freely available for all major platforms (Windows, Mac OS, and Linux).

```{r, eval=FALSE}
# Export trajectory graph structure to graphml
# Color and label nodes by state
write.ygraphml(sce=exBundle,
               file='yourFileName.graphml',
               color_by='phenoName',
               name='state',
               node_label='state')
```

Let's open the exported graphml file in _yEd_:

```{r, eval=TRUE, echo=FALSE}
pth <- "img/yEd_1.gif" #system.file("img", "yEd_1.gif", package="CellTrails")
knitr::include_graphics(path=pth)
```

If a layout has already been defined for a `r Biocpkg("SingleCellExperiment")` object, the samples' coordinates will be listed in the exported graphml file and will be directly interpreted by _yEd_. In this example, no layout was defined yet and therefore, all samples (nodes) are overlapping.

*Please note* that, in this example, the export function `write.ygraphml` was set to colorize nodes by their assigned state. However, it is possible to colorize and label nodes by any phenotype or feature expression data stored in the `r Biocpkg("SingleCellExperiment")` object. For example, we could colorize the nodes by the recorded FM1-43 dye intensity to get an idea where the trajectory might start and end (a high FM1-43 dye indicates mature cells) and label the nodes by their determined state by setting the parameters `color_by="phenoName"`, `name="fm143"` and `label="state"`. Nodes with a missing phenotype information are not colorized and remain transparent.

Next, we layout the graph. Since the trajectory resembles a tree structre, we use a tree algorithm. We found that routing the trajectory graph in a quasi-radial style, called balloon style, works very well for our case. The layouter can be selected via  _Layout_ $\rightarrow$ _Tree_ $\rightarrow$ _Balloon_. The following parameter setting was used in the original _CellTrails_ publication:

```{r, eval=TRUE, echo=FALSE}
pth <- "img/yEd_2-3.gif" #system.file("img", "yEd_2-3.gif", package="CellTrails")
knitr::include_graphics(path=pth)
```

Let's run the algorithm to compute the layout:

```{r, eval=TRUE, echo=FALSE}
pth <- "img/yEd_4.gif" #system.file("img", "yEd_4.gif", package="CellTrails")
knitr::include_graphics(path=pth)
```

*Please note* that edge crossings (i.e., two or more edges overlap) are not useful and if they occur we suggest to re-run the layouter with different parameters before saving the layout. Using either your mouse or the _View_ $\rightarrow$ _Zoom In_ option allows to have a closer look. If we want to have the trajectory progressing from bottom left to bottom right (based on a specific feature expression or phenotype label), we need to transform the graph. This can be done via _Tools_ $\rightarrow$ _Geometric transformations_. Here, we select _Mirror on Y axis_ and _Mirror on X axis_.

```{r, eval=TRUE, echo=FALSE}
pth <- "img/yEd_5-6.gif" #system.file("img", "yEd_5-6.gif", package="CellTrails")
knitr::include_graphics(path=pth)
```

Finally, the file can be saved via _File_ $\rightarrow$ _Save_ and reimported to R. 

```{r, eval=FALSE}
tl <- read.ygraphml("yourFileName.graphml")
```

For illustration purposes, the computed trajectory layout for the example dataset is available within this package.

```{r, eval=TRUE, echo=FALSE}
par(mar=c(0,0,0,0))
```

```{r, eval=TRUE}
# Import layout
tl <- read.ygraphml(system.file("exdata", "bundle.graphml", 
                                package="CellTrails"))

# Plot layout
plot(tl[,1:2], axes=FALSE, xlab="", ylab="", pch=20, cex=.25)
```

Finally, we set the trajectory layout to the `r Biocpkg("SingleCellExperiment")` object using the `trajLayout` function. Here, the parameter `adjust` indicates if edge lengths should be adjusted, such that they correspond to the inferred pseudotime.
```{r, eval=TRUE}
# Adjust layout and store to object
trajLayout(exBundle, adjust=TRUE) <- tl

showTrajInfo(exBundle)

# Plot adjusted layout
plot(trajLayout(exBundle), axes=FALSE, xlab="", ylab="", pch=20, cex=.25)
```

**Sidenote**. It is not a requirement to use yEd. The trajectory layout can also be defined by the user otherwise. The minimum requirement is, that the coordinates of each sample are stored in a `data.frame` whose row names correspond to the trajectory sample names. The sample names can be pulled from the `r Biocpkg("SingleCellExperiment")` object using the function `trajSampleNames`. The layout can then be set using the accessor function `trajLayout` accordingly.

## Plot Maps
After generating the layout, a two-dimensional visualization of the trajectory can be drawn. Here, the line represents the trajectory and individual points represent the samples. This plot type either colorizes individual samples by a metadata label (as available via `phenoNames`) or it shows the topography of a given feature's expression landscape (as available via `featureNames`). When metadata are being visualized, the grey line represents the trajectory and the individual points represent samples. Samples that do not have a specific metadata label or a missing value are not shown. Let's visualize how the cellular FM1-43 uptake distributes along the trajectory:
```{r, eval=TRUE}
plotMap(exBundle, color_by="phenoName", name="fm143")
```

In the topographical plot, a smooth surface is fitted and values are predicted for a regular grid resulting in the shown map topography; the red line signifies the trajectory. Let's take a view into the _ATOH1_ expression landscape, an early key transcription factor during sensory hair cell development:

```{r, eval=TRUE}
plotMap(exBundle, color_by="featureName", name="ATOH1", type="surface.fit")
```

Let's have a look into the standard error of the predicted expression surface by setting the parameter `type`.
```{r, eval=TRUE}
plotMap(exBundle, color_by="featureName", name="ATOH1", type="surface.se")
```

Alternatively, *CellTrails* enables to show the samples only, instead of the whole fitted expression surface. Here, we have two options: either the raw expression or the smoothed values.

```{r, eval=TRUE}
# Raw
plotMap(exBundle, color_by="featureName", name="ATOH1", type="raw")

# Smoothed
plotMap(exBundle, color_by="featureName", name="ATOH1", type="surface.fit", 
        samples_only=TRUE)
```
